---
title: "GGG-201D-problem-set-2"
author: "Ethan Holleman"
date: "4/28/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

### Part A
Use a program such as R or Excel to generate a scatter plot that shows how expected allele frequency change from genetic drift depends on initial allele frequency. The x-axis should be initial allele frequency and range from 0 to 1. The y-axis should be expected change in allele frequency after one generation. Perform calculations in steps of 0.1 for a population size of 2N=20.

```{r}

expected <- function(init_allele_freq, pop_size, delta_allele_freq){
  
  num_delta_alleles <- pop_size * delta_allele_freq
  num_init_alleles <- pop_size * init_allele_freq
  routes <- num_init_alleles + c(-1*num_delta_alleles, num_delta_alleles)
  routes <- unique(routes[routes >= 0])  # remove negative values since not possible
  sum(mapply(dbinom, routes, pop_size, init_allele_freq))
  
}
```

```{r}
expected_change <- function(init_allele_freq, pop_size){
  
  magnitudes <- seq(0, 1-init_allele_freq, 0.1)
  sum(mapply(expected, init_allele_freq, pop_size, magnitudes) * magnitudes)

}
```


```{r}
library(ggplot2)
library(ggpubr)

df.a <- data.frame(init_allele_freq=seq(0, 1, 0.1), pop_size=10)
df.a$delta_allele_freq <- mapply(expected_change, df.a$init_allele_freq, df.a$pop_size)
ggplot(df.a, aes(x=init_allele_freq, y=delta_allele_freq)) + geom_line(color='dodgerblue') +
      theme_pubclean() + labs(x='Initial allele frequency', y='Change allele freq after 1 gen')
  
```

```{r, echo=FALSE}
colnames(df.a) <- c('Initial allele frequency', 'Pop size', 'Delta allele freq 1 gen')
knitr::kable(df.a)
```

### Part B

Use the same program to generate a scatter plot that shows how expected allele frequency change from genetic drift depends on population size. The x-axis should be population size and range from 2N=10 to 2N=100. The y-axis should be expected change in allele frequency after one generation. Perform calculations in steps of 10 with an allele frequency of 0.5

```{r}
df.b <- data.frame(init_allele_freq=0.5, pop_size=seq(10, 100, 10))
df.b$delta_allele_freq <- mapply(expected_change, df.b$init_allele_freq, df.b$pop_size)
ggplot(df.b, aes(x=pop_size, y=delta_allele_freq)) + geom_line(color='darkred') +
      theme_pubclean() + labs(x='Population size', y='Change allele freq after 1 gen')
```

```{r, echo=FALSE}
colnames(df.b) <- c('Initial allele frequency', 'Pop size', 'Delta allele freq 1 gen')
knitr::kable(df.b)
```

## Question 2

You sequence a locus in three individuals from a population and obtain the below data:
```
I1:
  GCTACTTTACCATTCTCAGCGAGACGTAAGATCAGGCCAGATCCACCTCG
  GTTCCTTTAACATTCTCAGCGAGACGTAAGAGCAGGCCAGATCCACCGCC
I2:
  GCTCCTTTACCATCCTCAGCGAGACGTAAGATCAGGACAGATCCACCTCG
  GCTACTTTACCATCCTCAGCGACACGTAAGATCAGGCCAGATCCACCTCG
I3:
  GCTACTTTACCATCCTCAGCGAGACGTAAGAGCAGGCCAGATCCACCTCC
  GCTACTTTACCATCCTCAGCGAGACGTAGGAGCAGGACAGATCCACCTCG
```

### Part A

How many segregating sites (s) are present in these data?

```{r}

seq.df <- data.frame(
  i1.1=unlist(strsplit('GCTACTTTACCATTCTCAGCGAGACGTAAGATCAGGCCAGATCCACCTCG', '')),
  i1.2=unlist(strsplit('GTTCCTTTAACATTCTCAGCGAGACGTAAGAGCAGGCCAGATCCACCGCC', '')),
  i2.1=unlist(strsplit('GCTCCTTTACCATCCTCAGCGAGACGTAAGATCAGGACAGATCCACCTCG', '')),
  i2.2=unlist(strsplit('GCTACTTTACCATCCTCAGCGACACGTAAGATCAGGCCAGATCCACCTCG', '')),
  i3.1=unlist(strsplit('GCTACTTTACCATCCTCAGCGAGACGTAAGAGCAGGCCAGATCCACCTCC', '')),
  i3.2=unlist(strsplit('GCTACTTTACCATCCTCAGCGAGACGTAGGAGCAGGACAGATCCACCTCG', ''))
)

site_is_segregating <- function(row){
  
  copy1 <- length(unique(unlist(row[seq(1, length(row), 2)]))) == 1
  copy2 <- length(unique(unlist(row[seq(2, length(row), 2)]))) == 1
  if (copy1 & copy2){
    FALSE
  }else{
    TRUE
  }
}

seg_sites <- sum(apply(seq.df, 1, site_is_segregating))

message(paste('Number of segregating sites:', seg_sites))
```
### Part B

What is pi in these data?

```{r}

pi <- function(seq.df){
  
  comparisons <- combn(1:ncol(seq.df), 2)
  distances <- c()
  for (i in 1:ncol(comparisons)){
    col_a <- comparisons[1, i]
    col_b <- comparisons[2, i]
    num_diffs <- nrow(seq.df) - (sum(seq.df[, col_a] == seq.df[, col_b]))
    distances <- c(distances, num_diffs)
  }

 sum(distances) / ncol(comparisons)
}

pi_genes <- pi(seq.df)
message(paste('Pi =', pi_genes))
```

### Part C

What are s and pi expressed in per site values?

```{r}
message(paste('Pi per site =', pi_genes / nrow(seq.df)))
message(paste('Number of segregating sites per site:', seg_sites / nrow(seq.df)))
```

### Part D

What is the minor allele frequency spectrum for these data?

```{r}
# number of sites with given minor allele freq, max could be 3 / 6 otherwise
# major allele
# number sites is length of seq

minor_allele_count <- function(row){
  
  # get number of unique alleles need to determine what major / minor allele is
  # if 1 then only 1 allele so minor allele = 09
  row.ul <- as.vector(unlist(row))
  unq.alleles <- unique(row.ul)
  if (length(unq.alleles) == 1){
    0
  }else{
    # should be length 2 if not 1
    allele.a.count <- sum(row.ul == unq.alleles[1])
    allele.b.count <- sum(row.ul == unq.alleles[2])
    # minor allele freq will be the min of these values
    min(allele.a.count, allele.b.count)
  }
}

```

```{r}
seq.df.minor <- seq.df
seq.df.minor$minor <- apply(seq.df, 1, minor_allele_count)

ggplot(seq.df.minor, aes(x=minor)) + geom_bar(fill='dodgerblue', color='black') + 
  theme_pubclean() + labs(x='Minor allele frequency', y='Number of sites')

```

### Part E

You next sequence the locus in a few closely related species and determine the ancestral sequence to bethe following.

```
Ancestral: GCTCCTTTACCATCCTCAGGGACACGTAAGAGCAGGCCAGACCCACCTCC
```

What is the derived allele frequency spectrum for these data?

```{r}
# add ancestral seq to df
ancestoral <- 'GCTCCTTTACCATCCTCAGGGACACGTAAGAGCAGGCCAGACCCACCTCC'
seq.df.anc <- seq.df
seq.df.anc$ancestor <- unlist(strsplit(ancestoral, ''))
```

```{r}
derived_allele_freq <- function(row){
  
  # assume ancestral allele is last value
  row.ul <- as.vector(unlist(row))
  ansestoral <- row[length(row)]
  sum(row[1:length(row)-1] != ansestoral)
  
}
```

```{r}

derived <- apply(seq.df.anc, 1, derived_allele_freq)
ggplot(as.data.frame(derived), aes(x=derived))  + geom_bar(fill='tan', color='black') + 
  theme_pubclean() + labs(x='Derived allele frequency', y='Number of sites')
```

## Question 3

Use a program such as R or Excel to generate a scatter plot that shows the properties of the coalescent process in a Wright-Fisher population. The x-axis should be number of gene copies and range from 2-50. The y-axis should be expected number of generations in N units. Perform calculations in steps of one gene copy and plot the following three expectations: (1) time to the first coalescent event; (2) time to the most recent common ancestor of all gene copies; (3) total tree length.
```{r}
gene_copies <- 2:50
```

```{r}
co.df <- data.frame(gene_copies=gene_copies)
# E(time to co event ) = 2n / number of gene copies
first_co_event <- function(copies){
  2 / copies
}

co.df$first_co_event <- unlist(lapply(gene_copies, first_co_event))
```

```{r}

tmrca <- function(gene_copies){
  
  vals <- c()
  for (k in 2:gene_copies){
    vals <- c(vals, (2 / ((k*(k-1)) / 2)))
  }
  sum(vals)
}

co.df$tmrca <- unlist(lapply(gene_copies, tmrca))

```

```{r}

ttl <- function(gene_copies){
   vals <- c()
  for (k in 2:gene_copies){
    vals <- c(vals, (2 / ((k*(k-1)) / 2)) * k)
  }
  sum(vals)
}

co.df$ttl <- unlist(lapply(gene_copies, ttl))

```

```{r, warning=FALSE}
library(reshape2)
library(ggsci)
co.df.melt <- melt(co.df, id.vars="gene_copies")
ggplot(co.df.melt, aes(x=gene_copies, y=value, color=variable)) +
  geom_line() + theme_pubclean() + 
  labs(y='Expected number generations (N)', x='Gene copies') +
  scale_color_discrete(name = "variable", 
                      labels = c("First coalenscent event", "tMRCA", "TTL"),
                      ) +
  scale_color_futurama()

```
